Merging merges, more or less

### 2015-03-29 2015mergingmerges/section/01-Introduction.tex
little things

Fusion for array programs is important for removing intermediate arrays, reducing memory traffic as well as allocations.
However, when dealing with data too large to fit in memory such as tables on disk, removing intermediate arrays becomes essential rather than just desirable.
Attempting to create an intermediate array of such amounts of data would lead to thrashing and swapping to disk, or perhaps even running out of swap.
For these situations, some sort of assurance of total fusion are required: either the program can be fused with no intermediate arrays or unbounded buffers, or it will not compile at all.

@@

Sadly, this program requires buffering the entire @xs@ source in memory while performing the first actions, then looping through the buffered @xs@ and performing the second action afterwards.
However, we could execute this with no buffering if the relative ordering of the two @mapM_@s was not important, and could be reordered and interspersed as the data is read.
We use the following @When@ syntax to denote performing a reorderable IO action \emph{when}ever data is available.

@@

Sadly this last program requires an unbounded buffer or two iterations over @xs@, so we would like to disallow it.
It may not be immediately obvious just looking at the program that it requires two iterations, which is why we would like this to be automated.

>>>>

Fusion for array programs is important for removing intermediate arrays, reducing memory traffic as well as allocations.
However, when dealing with data too large to fit in memory such as tables on disk, removing intermediate arrays becomes essential rather than just desirable.
Attempting to create an intermediate array of such amounts of data would lead to thrashing and swapping to disk, or perhaps even running out of swap.
For these situations, some sort of assurance of total fusion is required: either the program can be fused with no intermediate arrays or unbounded buffers, or it will not compile at all.

@@

Sadly, this program requires buffering the entire @xs@ source in memory while performing the first actions, then looping through the buffered @xs@ and performing the second action afterwards.
However, we could execute this with no buffering if the relative ordering of the two @mapM_@s were not important, and could be reordered and interspersed as the data is read.
We use the following @When@ syntax to denote performing a reorderable IO action \emph{when}ever data is available.

@@

Sadly this last program requires an unbounded buffer or two iterations over @xs@, so we would like to disallow it.
It may not be immediately obvious just looking at this program alongside the earlier ones that it requires two iterations, which is why we would like this to be automated.

### 2015-03-29 2015mergingmerges/section/02-Machines.tex
little things

We describe a restricted automaton as a eight-tuple, extending the standard description by three elements: $(Q, \Sigma, \delta, q_0, F, \Gamma, \gamma, \sigma)$.

@@

\[ \delta : \forall q : Q. \sigma(\gamma(q)) \to Q \]
For each state, the transition function need only be defined for those letters of the state's type.

@@

\subsection{State types}
We now describe a set of types and alphabet for describing combinator programs as a restricted DFA.
These are parameterised by both the names of sources, sinks and bindings $n$, and the type of worker functions $f$ (such as @Expression@ for code generation).

@@

\subsection{Invariants}
Not all machines are valid or meaningful, and we wish to rule them out.
For example, a machine with an initial state of $@If@~(x>5)~n$ is meaningless because there is no local $x$ that has been read from an input. 
On the other hand, some programs could be meaningful, but requiring them to be in a particular form simplifies fusion, later.
An example of such is a program that does not release its input before pulling again; a sympathetic code extraction would be to simply release before every pull, but the explicit pulls makes synchronising two machines easier.

@@

To begin, we annotate the initial state with an empty set of information: at the start of the machine, we have no read values available, nor is it known (or likely!) that any channels are finished.
Then, for each output transition, we annotate the output state according to the current state type and the transition label: @Pull n@ adds a @Value n@ or @Finished n@ to its @Some n@ and @None@ transitions.
If the output state has already been annotated, we check that the two annotations are the same, modulo values produced by @Out@.
The reason that annotations may be different in the values produced by @Out@ is that the @Out@ values are not explicitly released, and may be referred to by other functions, for example after fusing two combinators together.
\input{figures/Invariants.tex}

After each state is annotated, we check that its annotation allows the state type 
For each state, we also check that its annotation allows its execution: a @Pull n@ cannot run if $n$ is already finished; @Release n@ can only run if there is a @Value n@ to release, and so on.

@@

@@

For example, the @filter@ in Figure~\ref{fig:com:filter} is compiled to something like the following:
\begin{code}
run :: IO Int -> (Maybe Int -> IO ()) -> IO ()
run pull_xs when_o = l10
 where
  l10 = do  x  <- pull_xs
            case x of
             Nothing -> l90
             Just x' -> l15 x'

@@

  l91   =   return ()
\end{code}

>>>>

We describe a restricted automaton as an eight-tuple, extending the standard description by three elements: $(Q, \Sigma, \delta, q_0, F, \Gamma, \gamma, \sigma)$.

@@

\[ \delta ~:~\forall q : Q.\ \sigma(\gamma(q))~\to~Q \]
For each state, the transition function need only be defined for those letters of the state's type.

@@

\subsection{State types}
We now describe the set of types and alphabet for describing combinator programs as a restricted DFA.
These are parameterised by both the names of sources, sinks and bindings $n$, and the type of worker functions $f$ (such as @Expression@ for code generation).

@@

\subsection{Invariants}
Not all machines are valid or meaningful, and we wish to rule them out.
For example, a machine with an initial state of $@If@~(x>5)~n$ is meaningless because there is no local $x$ that has been read from an input. 
On the other hand, some programs could be meaningful, but requiring them to be in a particular form simplifies fusion, later.
An example of such is a program that does not release its input before pulling again; a sympathetic code extraction would be to simply release before every pull, but having explicit releases makes synchronising two machines easier.

@@

To begin, we annotate the initial state with an empty set of information: at the start of the machine, we have no read values available, nor is it known (or likely!) that any channels are finished.
Then, for each output transition, we annotate the output state according to the current state type and the transition label: @Pull n@ adds a @Value n@ or @Finished n@ to its @Some n@ and @None@ transitions respectively.
If the output state has already been annotated, we check that the two annotations are the same, modulo values produced by @Out@.
The reason that annotations may be different in the values produced by @Out@ is that the @Out@ values are not explicitly released, and may be referred to by other functions, for example after fusing two combinators together.
These @Out@ values are then implicitly released.

For each state, we check that its annotation allows its execution: a @Pull n@ cannot run if $n$ is already finished; @Release n@ can only run if there is a @Value n@ to release, functions of @Out@ and @If@ require their values to be present, and so on.

\input{figures/Invariants.tex}

@@

$$ \fvs{f} \subset a $$
Any free variables mentioned in a function must exist as @Value@s in the annotation, in order for the function to use these values.

@@

For example, the @filter@ in Figure~\ref{fig:com:filter} is compiled to something like the Figure~\ref{fig:extract:filter}.

\begin{figure}
\begin{code}
run :: IO Int -> (Maybe Int -> IO ()) -> IO ()
run pull_xs when_o = l10
 where
  l10 = do  x  <- pull_xs
            case x of
             Nothing -> l90
             Just x' -> l15 x'

@@

  l91   =   return ()
\end{code}
\caption{Extracted code for filter}
\label{fig:extract:filter}
\end{figure}

### 2015-03-29 2015mergingmerges/section/03-Access.tex
start on merging inference rules
working on invariants
rules for general merge
little things

\begin{tabbing}
MM \= MM \= MMMMMM \= M\kill
$\Psi$ \> $=$  \> @Value@     \> $n$         \\
       \> $~|$ \> @Finished@  \> $n$         \\
\end{tabbing}

The @merge@ function takes two machines with states $l_1$ and $l_2$, and produces a new machine with state $l_1 \times l_2 \times \{\Psi\}$.
It starts by creating an empty machine, and inserting states one by one, starting from the initial state as the initial state of each machine, and empty pending sets.
Then, for each output transition of the input machines, states are recursively added.

\begin{tabbing}
MMMMMMM \= MM \= MMMMMM \= M\kill
@merge@ \> $:$ \> $@Machine @l_1 \times @Machine @l_2$ \\
        \> $\to$ \> $@Machine @(l_1 \times l_2 \times \{\Psi\})$ \\
\\
@merge @$m_1~m_2$ \> $=$ \> $@merge@'~(@init @m_1)~(@init @m_2)~\{\}~@empty@$ \\
\\
$@merge@'$ \> $:$ \> $l_1 \times l_2 \times \{\Psi\} \times @Machine @(l_1 \times l_2 \times \{\Psi\})$ \\
           \> $\to$ \> $@Machine @(l_1 \times l_2 \times \{\Psi\})$ \\
\\
MMM \= M \= MMMMMMMMMM \= M \=\kill
$@merge@'~s_1~s_2~\psi~m$ \\
 \> $~|$ \> $(s_1, s_2, \psi) \in m$  \\
 \> $=$  \> $m$ \\
 \> $~|$ \> $(\gamma,\delta) \in @move@~s_1~s_2~\psi$ \\
 \> $=$  \> $@fold@~@merge@'~(m \cup \gamma \cup \delta)~\delta$ \\
\end{tabbing}

@@

\begin{tabbing}
MMM \= M \= MMMMMMM\kill
$@move@$ \> $:$ \> $l_1 \times l_2 \times \{\Psi\}$ \\
           \> $\to$ \> $\Gamma \times \{\Sigma \times (l_1 \times l_2 \times \{\Psi\})\}$ \\
\end{tabbing}

@@

\begin{tabbing}
MMM \= M \= MMMMMMM \= M \=\kill
 \> $~|$ \> $@Update@~f~n$ \> $=$ \> $@state@~m_1~s_1$ \\
 \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
 \> $=$ \> $(@Update@~f~n,~\{(@Unit@, (s_1',s_2,\psi))\})$ \\
\\
 \> $~|$ \> $@Update@~f~n$ \> $=$ \> $@state@~m_2~s_2$ \\
 \> $\wedge$ \> $s_2'$     \> $=$ \> $@trans@~m_2~s_2~@Unit@$ \\
 \> $=$ \> $(@Update@~f~n,~\{(@Unit@, (s_1,s_2',\psi))\})$ \\
\end{tabbing}

@@

\begin{tabbing}
MMM \= M \= MMMMMMM \= M \=\kill
 \> $~|$ \> $@Skip@$ \> $=$ \> $@state@~m_1~s_1$ \\
 \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
 \> $=$ \> $(@Skip@,~\{(@Unit@, (s_1',s_2,\psi))\})$ \\
\\
 \> $~|$ \> $@Skip@$ \> $=$ \> $@state@~m_2~s_2$ \\
 \> $\wedge$ \> $s_2'$     \> $=$ \> $@trans@~m_2~s_2~@Unit@$ \\
 \> $=$ \> $(@Skip@,~\{(@Unit@, (s_1,s_2',\psi))\})$ \\
\end{tabbing}

@@

\begin{tabbing}
MMM \= M \= MMMMMMM \= M \=\kill
 \> $~|$ \> $@If@~f~n$ \> $=$ \> $@state@~m_1~s_1$ \\
 \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@True@$ \\
 \> $\wedge$ \> $s_1''$     \> $=$ \> $@trans@~m_1~s_1~@False@$ \\
 \> $=$ \> $(@If@~f~n,~\{(@True@, (s_1',s_2,\psi)), (@False@, (s_1'', s_2, \psi))\})$ \\
\\
 \> $~|$ \> $@If@~f~n$ \> $=$ \> $@state@~m_2~s_2$ \\
 \> $\wedge$ \> $s_2'$     \> $=$ \> $@trans@~m_2~s_2~@True@$ \\
 \> $\wedge$ \> $s_2''$     \> $=$ \> $@trans@~m_2~s_2~@False@$ \\
 \> $=$ \> $(@If@~f~n,~\{(@True@, (s_1,s_2',\psi)), (@False@, (s_1, s_2'', \psi))\})$ \\
\end{tabbing}

@@

\begin{tabbing}
MMM \= M \= MMMMMMM \= M \=\kill
 \> $~|$ \> $@Done@$ \> $=$ \> $@state@~m_1~s_1$ \\
 \> $\wedge$ \> $@Done@$ \> $=$ \> $@state@~m_2~s_2$ \\
 \> $=$ \> $(@Done@,~\{\})$ \\
\end{tabbing}

@@

\begin{tabbing}
MMM \= M \= MMMMMMM \= M \=\kill
 \> $~|$ \> $@Out@~f~n$ \> $=$ \> $@state@~m_1~s_1$ \\
 \> $\wedge$ \> $n$      \> $\in$ \> $@inputs@~m_2$ \\
 \> $\wedge$ \> $@Done@$ \> $=$ \> $@state@~m_2~s_2$ \\
 \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
 \> $=$ \> $(@Out@~f~n,~\{(@Unit@, (s_1',s_2,\psi \setminus \{@Value@~n\} ))\})$ \\
\end{tabbing}

If the first machine is producing an output on channel $n$ and the second machine uses $n$, the first machine would usually have to wait until the second machine is ready to accept new data.
However, if the second machine is @Done@, the first machine may output to the channel as frequently as it likes.
Note that the output channel $n$ may have multiple consumers, so continuing to output after the second machine consumer has finished is not as futile as it may seem at first.
\TODO{Checking for @Done@ of machine is too coarse-grained, need to check if $n$ is closed instead.}

\begin{tabbing}
MMM \= M \= MMMMMMM \= M \=\kill
 \> $~|$ \> $@Out@~f~n$ \> $=$ \> $@state@~m_1~s_1$ \\
 \> $\wedge$ \> $n$      \> $\in$ \> $@inputs@~m_2$ \\
 \> $\wedge$ \> $@Value@~n$ \> $\not\in$ \> $\psi$ \\
 \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
 \> $=$ \> $(@Out@~f~n,~\{(@Unit@, (s_1',s_2,\psi \cup \{@Value@~n\} ))\})$ \\
\end{tabbing}

Again, if the first machine is producing and the second is consuming, the first machine may only produce if there is no unhandled value on the channel.
This is what $\psi$ is for.
If there is a value in $\psi$, a later case may allow the second machine to pull from it.

\begin{tabbing}
MMM \= M \= MMMMMMM \= M \=\kill
 \> $~|$ \> $@Out@~f~n$ \> $=$ \> $@state@~m_1~s_1$ \\
 \> $\wedge$ \> $n$      \> $\not\in$ \> $@inputs@~m_2$ \\
 \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
 \> $=$ \> $(@Out@~f~n,~\{(@Unit@, (s_1',s_2,\psi))\})$ \\
\end{tabbing}

The final case for @Out@ is when the second machine does not use this output, allowing the first machine to output at any time.
The cases where the second machine is on an @Out@ are symmetric to these.

The cases for @OutDone@ proceed similarly, and are also symmetric for the first and second machines.

\begin{tabbing}
MMM \= M \= MMMMMMM \= M \=\kill
 \> $~|$ \> $@OutDone@~n$ \> $=$ \> $@state@~m_1~s_1$ \\
 \> $\wedge$ \> $n$      \> $\in$ \> $@inputs@~m_2$ \\
 \> $\wedge$ \> $@Done@$ \> $=$ \> $@state@~m_2~s_2$ \\
 \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
 \> $=$ \> $(@OutDone@~n,~\{(@Unit@, (s_1',s_2,\psi))\})$ \\
 \\
 \> $~|$ \> $@OutDone@~n$ \> $=$ \> $@state@~m_1~s_1$ \\
 \> $\wedge$ \> $n$      \> $\in$ \> $@inputs@~m_2$ \\
 \> $\wedge$ \> $@Value@~n$ \> $\not\in$ \> $\psi$ \\
 \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
 \> $=$ \> $(@OutDone@~n,~\{(@Unit@, (s_1',s_2,\psi \cup \{@Finished@~n\}))\})$ \\
 \\
 \> $~|$ \> $@OutDone@~n$ \> $=$ \> $@state@~m_1~s_1$ \\
 \> $\wedge$ \> $n$      \> $\not\in$ \> $@inputs@~m_2$ \\
 \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
 \> $=$ \> $(@OutDone@~n,~\{(@Unit@, (s_1',s_2,\psi))\})$ \\
\end{tabbing}

@@

>>>>

\begin{tabbing}
MM \= MM \= MMMMMM \= M\kill
$\Psi$ \> $=$  \> @Value@     \> $n$         \\
       \> $~|$ \> @Finished@  \> $n$         \\
       \> $~|$ \> @Closed@    \> $n$         \\
\end{tabbing}

The @merge@ function takes two machines with states $l_1$ and $l_2$, and produces a new machine with state $l_1 \times \{\Psi\} \times l_2 \times \{\Psi\}$.
It starts by creating an empty machine, and inserting states one by one, starting from the initial state as the initial state of each machine, and empty pending sets.
Then, for each output transition of the input machines, states are recursively added.

\begin{tabbing}
MMMMMMM \= MM \= MMMMMM \= M\kill
@merge@ \> $:$ \> $@Machine @l_1 \times @Machine @l_2$ \\
        \> $\to$ \> $@Machine @(l_1 \times \{\Psi\} \times l_2 \times \{\Psi\})$ \\
\\
@merge @$m_1~m_2$ \> $=$ \> $@merge@'~(@init @m_1)~(@init @m_2)~\{\}~@empty@$ \\
\\
$@merge@'$ \> $:$ \> $l_1 \times \{\Psi\} \times l_2 \times \{\Psi\}$               \\
           \> $\times$ \> $@Machine @(l_1 \times \{\Psi\} \times l_2 \times \{\Psi\})$ \\
           \> $\to$ \> $@Machine @(l_1 \times \{\Psi\} \times l_2 \times \{\Psi\})$ \\
\\
MMM \= M \= MMMMMMMMMM \= M \=\kill
$@merge@'~s_1~\psi_1~s_2~\psi_2~m$ \\
 \> $~|$ \> $(s_1, \psi_1, s_2, \psi_2) \in m$  \\
 \> $=$  \> $m$ \\
 \> $~|$ \> $(\gamma,\delta) \in @move@~s_1~\psi_1~s_2~\psi_2$ \\
 \> $=$  \> $@fold@~@merge@'~(m \cup \gamma \cup \delta)~\delta$ \\
\end{tabbing}

@@

\begin{tabbing}
MMM \= M \= MMMMMMM\kill
$@move@$ \> $:$ \> $l_1 \times \{\Psi\} \times l_2 \times \{\Psi\}$ \\
           \> $\to$ \> $\Gamma \times \{\Sigma \times (l_1 \times \{\Psi\} \times l_2 \times \{\Psi\})\}$ \\
\end{tabbing}

@@

See figure~\ref{fig:merge:general}.

\input{figures/MergeGen.tex}

@@

@@

@@

@@

If the first machine is producing an output on channel $n$ and the second machine uses $n$, the first machine would usually have to wait until the second machine is ready to accept new data.
However, if the second machine has closed $n$, the will never pull on $n$ again, so the first machine may output to the channel as frequently as it likes.
Note that the output channels may have multiple consumers, so continuing to output after a consumer has finished is not as futile as it may seem at first.

Again, if the first machine is producing and the second is consuming, the first machine may only produce if there is no unhandled value on the channel.
This is what $\psi$ and $\phi$ are for.
If there is a value in the other machine's event set, $\phi$, a later case may allow the second machine to pull from it.

The final case for @Out@ is when the second machine does not use this output, allowing the first machine to output at any time.

The cases for @OutDone@ proceed similarly to @Out@.

@@

\subsection{Theorems we want to prove}
\subsubsection{General preservation}
$$
\forall a~b.\ a~@ok@~\wedge~b~@ok@~\wedge \forall c.\ @merge@~a~b = c \implies c~@ok@
$$

\subsubsection{Vertical preservation}
$$
\forall a~b.\ a~@ok@~\wedge~b~@ok@~\wedge \forall c.\ @mergeV@~a~b = c \implies c~@ok@
$$

\subsubsection{Vertical progress}
\begin{tabbing}
MM \= M \= \kill
$ \forall a~b.\ $ \> \>
$a~@ok@~\wedge~b~@ok@$ \\
\> $\wedge$ \> $@inputs@~a~\cap~@inputs@~b=\emptyset~\wedge~@outputs@~a~\cap~@outputs@~b=\emptyset$ \\
\> $\wedge~($ \> $\exists n.\ (@inputs@~a~\cap~@outputs@~b)~\cup~(@outputs@~a~\cap~@inputs@~b) = \sgl{n})$ \\
 $\implies$ \> \> $\exists c.\ @mergeV@~a~b=c$
\end{tabbing}
